---
title: "Assignment 5"
author: "hp621@scaretletmail.rutgers.edu"
date: "10/28/2022"
output: html_document
---
```{r}
#Importing Data set and view data set

library(readxl)
HH_NatrualGAS <- read_excel("C:/Users/HARSH/Desktop/RU/BF/assignment4/HH_NatrualGAS.xlsx", 
    col_types = c("date", "numeric"))
View(HH_NatrualGAS)
```

```{r}
#we use these libraries for Decomposition
library(fpp)
library(fpp2)
library(UsingR)
library(xts)
library(TTR) 
```
```{r}
#Check your dataset in time series
head(HH_NatrualGAS)
plot(HH_NatrualGAS)
ts_data <- ts(HH_NatrualGAS$Price, start = c(1997, 01),end = c(2020,12), frequency = 12)
summary(HH_NatrualGAS)

```
```{r}
#perform seasonal decomposition STL
stl_decomp <- stl(ts_data,s.window ="periodic")
#plot the decomposed components
plot(stl_decomp)
#print attributes of the decomposition
attributes(stl_decomp)
```

```{r}
# Lets print out a seasonal adjustment
seasadj(stl_decomp)
# Plot a line on the graph
plot(ts_data)
lines(seasadj(stl_decomp), col="Red")


```
```{r}
# Default period forecast
f_stl <- forecast(stl_decomp)
# you can pass the # of period
f_stl <- forecast(stl_decomp,h=15)
f_stl
plot(f_stl)
```

```{r}
# There is more than one way to do things
decomp_ts_data <- decompose(ts_data)

# Each one shows different attributes 
attributes(decomp_ts_data)
seasadj(decomp_ts_data)
```
####
Data Loading:
You first load the necessary libraries, read the time series data from an Excel file, and view the data.

Check Time Series Data:
You examine the first few rows of the data and create a time series object (ts_data) with a specified start and end date and a frequency of 12 (monthly data). You also print a summary of the dataset.

Seasonal Decomposition (STL):
You perform seasonal decomposition using the stl function. The stl_decomp object contains the decomposed components of the time series, including seasonal, trend, and remainder components. You plot the decomposed components for visualization.

Seasonal Adjustment:
You print the seasonal adjustment using the seasadj function on the stl_decomp. Then, you overlay the seasonal adjustment on the original time series plot.

Forecasting:
You generate forecasts using the forecast function on the stl_decomp. The first forecast is created with the default number of periods, and the second one is created with a specific number of periods (15 in this case). You also plot the forecasted values.

Alternative Decomposition:
You perform another decomposition using the decompose function on the ts_data. This results in a different decomposition, and you check the attributes of the decomposed data and print the seasonal adjustment.
